An Extension of LTL with Rules\\ and its Application to Runtime Verification

### 2019-02-05 rule-ltl.tex
fixed figures
fixed figs
Updates from Overleaf

\begin{figure}
\begin{center}
\includegraphics[height=1.1in,width=2.7in]{PROP.jpg}
\caption{\label{prop} An arrow from A to B means that logic  B is more expressive than A. A double headed arrow shows same expressiveness}
\end{center}
\end{figure}

@@

The formulas of the core FLTL logic are 
defined by the following grammar,
where $p$ denotes a relation $R$,
$a$ is a constant representing a value in $d ( R )$ and $x$ is a variable with $\domain ( x ) = d ( p )$.
\begin{center}
$\varphi ::= \true  \; | \;
    p ( a ) \; | \;
    p ( x ) \; | \;
    ( \varphi \wedge \varphi ) \;  |   \;
   \neg \varphi \; | \;
   \bigcirc \varphi \; | \; 
   ( \varphi \; \until \; \varphi ) \; | $ %\\
   $ \ominus \; \varphi \; | \;
    ( \varphi  \; \since  \; \varphi ) \; | \;
    \exists x \; \varphi$
\end{center}

@@

Additional operators are defined as in the propositional logic. We שךדם define
$\forall x \; \varphi = \neg \exists x \neg \varphi$.
Restricting the modal operators to the past operators
($\since$, $\ominus$ and the ones obtained from them) 
forms the logic PFLTL.
The quantification over values of variables here is `static' in the sense that they are independent of the state in the execution. We
denote static quantification with $\exists$ and
$\forall$.
Dynamic quantification, as defined for LTL in Section~\ref{sec:extending-prop-ltl},
and will extend FLTL below, is expressed with $\Exists$ and $\Forall$.

A model is a set of temporal
 relations ${\cal R} = \{ R_1 \, \ldots , R_m \}$.
 Since the standard definition of temporal logic is
 over a sequence (``the execution''). let
 ${\cal R} [ i ] = \{ R_1 [ i ] \, \ldots , R_m [ i ] \}$. A model ${\cal R}$ can thus be
seen as the sequence ${\cal R} [ 1 ]\, {\cal R} [ 2 ] \ldots$.

@@

Let $\vars ( \varphi )$ be the set of free (i.e., unquantified) variables of a
subformula $\varphi$. 
%The interpretation of a subformula
%$\varphi$ over some finite sequence is as a set of assignments
%to the variables $\vars ( \varphi)$.
We denote by $\gamma |_{\vars ( \varphi )}$ the restriction (projection) of
an assignment $\gamma$ to the free variables appearing in $\varphi$.
%We are careful in combining the sets of assignments
%between different
%subformulas, e.g., through union and intersection (corresponding
%to disjunction and conjunction, respectively) and the quantification.
%This is done using the projection and extension functions over
%assignments defined above.
%
Let $\epsilon$ be the empty assignment. 
%
In any of the following cases, $( \gamma,  {\cal R} , i ) \models \varphi$
is defined when $\gamma$ is an
assignment over $\vars ( \varphi )$, and $i\ge 1$.

@@

\item $( \epsilon ,  {\cal R} , i )
 \models p_k ( a ) $ if $R_k ( a, i )$, where $a$ denotes a constant from  $d(p)$.

\item $( [ x \mapsto a ] , {\cal R} , i ) \models p_k ( x )$ if $R_k ( a , i )$.

@@

\item $( \gamma , {\cal R} , i ) \models ( \varphi \; \until \; \psi )$ if
for some $j$, $j > i$, $( \gamma |_{\vars ( \psi )} , {\cal R} , j) 
\models \psi $ and for each $k$,  $i \leq k < j $,
$( \gamma |_{\vars ( \varphi )} , {\cal R} , k) \models \varphi$.

@@

Then for an FLTL (PFLTL) formula with no free variables, ${\cal R} \models \varphi$ iff $( \epsilon , {\cal R} , 1 ) \models \varphi$.
We will henceforth sometimes abuse notation, as is very often done in logic, and use the same
symbols both for the relations (semantics) and their representation in
the logic (syntax).

@@

We first demonstrate that the lack of expressiveness
carries over from LTL (PLTL) to FLTL (PFLTL).

\noindent {\bf Example.}
Let $p$ and $q$ be temporal relations, where their
restrictions to the state $s_i$, $p[i]$ and $q[i]$, respectively, are unary relations $p(x)$ and $q(x)$. The
specification that we want to monitor is that for each value
$a$, $p(a)$ appears in all the states where $q(a)$ has appeared an even number of times so far (for the odd occurrences, $p(a)$ can also appear, but does not have to). This is an extension of
Wolper's example over LTL, carried over to the first-order case. To show that this is not expressible in FLTL (and PFLTL),
consider models (executions) where only one data element $a$ appears. We can prove that in this case, by a simple structural induction on (sub)formulas, that for each property $\varphi$
we can replace occurrences of variables within
predicates by $a$, i.e., $p(x)$ and $q(x)$ become $p(a)$
and $q  (a )$, respectively, and that we can throw away
the quantification ($\forall x$, $\exists x$), obtaining
an equivalent formula. This is then equivalent to
an LTL formula that is obtained by replacing
the occurrences of $p(a)$ and $q(a)$ by propositions
$p_a$ and $p_b$, respectively. However, by~\cite{Wolper},
such a formula cannot express the said property.
This example demonstrates that the so called ``non-counting''~\cite{Thomas} property of LTL carries over
to FLTL.

\vspace{1ex}
\noindent {\bf Example 2.}
Consider the property that asserts that
a main process $y$ (where
$mathit{main} (x)$ appears
in the sequence), is a descendant of a process $x$ if $x$ spawns a chain of processes spawning (of
the form $\mathit{spawn} (z, w$ that leadsfrom $x$ to $y$. This property cannot be expressed in FLTL. The reason
is that the required property needs to
assert about the transitive closure
of the relation $\mathit{spawn}$. 
FLTL can be translated in a way similar to a standard translation of $LTL$ to monadic first order logic~\cite{Thomas}. Here, the relations will be written with their explicit last ``time'' parameter, and the temporal operators are replaced with first order quantification.
E.g., $\Box \forall x \, p(x) \rightarrow q(y)$ will be translated into $\forall t \, \forall x \, (p(x, i) \rightarrow \exists t' p(x, j) )$
\footnote{The obtain model will have, in addition
to the aforementioned relations, also
the linear order $<$ over the naturals.}
It is possible to show that the transitive closure of $\mathit{spawn}$ cannot be expressed in this first order setting. The proof is based on the
compactness theory of first order logic~\cite{Flum}.

@@

Extending FLTL (PFLTL) with quantifiers over relations, we obtain QFLTL
(and the past-restricted version QPFLTL, respectively). 
The syntax includes $\Exists p \, \varphi$, 
where $p$ denotes an auxiliary relation.
We also allow $\Forall p\ \varphi = \neg \Exists \neg \varphi$. The semantics is as follows.

\klaus{Below, has the notation ${\cal R}' |_{\{ q \} }$ been defined?}

\begin{itemize}
\item $(\gamma ,  {\cal R}, i) \models \Exists q \, \varphi$ iff there exists
${\cal R}'$ such that ${\cal R}' |_{\{ q \} } = {\cal R}$ and 
$( \gamma, {\cal R}' , i) \models \varphi$.
\end{itemize}
Consequently, quantification over relations effectively extends the model ${\cal R}$ into a model ${\cal R}'$ within the scope of the quantifier. Note that quantification here is dynmaic (as in QLTL and QPLTL).

\subsubsection{ExtendingFLTL  with Rules.}

We now extend FLTL into EFLTL in a way that is motivated by the  propositional extension from LTL (PLTL) to ELTL (EPLTL). We thus allow the following formula:
\begin{equation}
\label{EFLTL}
\psi \mathit{\ where\ } r_j  ( x_j ) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} 
\end{equation}
\begin{enumerate}
\item $\psi$, the {\em statement}, is a FLTL formula with
no free variables, 
\item $\varphi_j$ are PFLTL formulas with a single
free variable $x_j$,
\item $r_j$ is an auxiliary temporal relation with $2$ parameters; the first parameter is of the same 
type as $x_j$ and the second one is, as usual,
a natural number that is omitted in the temporal formulas. 
An auxiliary relation $r_j$ can appear within $\psi$. It can also appear in $\varphi_k$ of a {\em rule} $r_k := \varphi_k$, but only within the
scope\footnote{This restriction has the same motivation as in
the propositional version ELTL (and EPLTL): to avoid making the
formula $\false$ just by means of the definition of the auxiliary relations.}
of a previous-time operator $\ominus$. 
\end{enumerate}

Again, it
is simple to extend the syntax and semantics to include
relations $r_j$ with different number of arguments, and, correspondingly, formulas $\varphi_j$ with corresponding number of free
variables\footnote{We chose not to do that here (and in the definition of the semantics of FLTL above), because it
requires multi-indexing notation that is trivial and adds no further insight. Our implementation indeed uses
this more general capability.}.
The logic EPFLTL is obtained by restricting the temporal modalities of EFLTL to the past ones:
$\since$ and $\ominus$, and those obtained from them.

\begin{lemma} \label{samesame}
The auxiliary temporal relations $r_1, \ldots r_n$ are
uniquely defined from ${\cal R}$.
\end{lemma}
\noindent {\bf Proof.} by a simple induction, as in Lemma~\ref{fourone}. \qed

The following formula expresses the property that was shown to be 
not expressible using FLTL, at the beginning of this section.
\begin{equation}
\Box \forall x \, (r(x)\rightarrow p(x)) \mathrm{\ where\ }
r(x) = ( q(x) \leftrightarrow \ominus \neg  r(x)) 
\label{eq:wolper-first-order}
\end{equation}

\noindent
We define\footnote{This is done for lack of space. Formal semantics 
can be given by constructing a set of temporal relations extended
with the auxiliary ones
inductively, as would be done in a detailed proof of Lemma~\ref{samesame}.} the semantics for the EFLTL 
specification (\ref{EFLTL}) 
by using the following equivalent
QFLTL formula:

@@

\noindent 
The analogous of Lemma~\ref{simplecase} shows that EPFLTL does not
need the past temporal operators besides $\ominus$.

\begin{lemma}
The full expressive power of EPFLTL can be obtained by allowing only 
specifications of the form
\[ \Box \psi \mathrm{\ where\ } r_j  ( x_j ) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} \]
where $\psi$ consists of a fully quantified single occurrence of some
auxiliary relation $r_j$.
\end{lemma}

\klaus{Something seems wrong in the above formulation.
E.g.: Is the $\Box$ a restriction?, what is ``fully quantified single occurrence of some auxiliary relation $r_j$''?}

@@

\begin{theorem}
\label{theo2}
The expressive power of EFLTL is stricktly weaker than that of QFLTL.
\end{theorem}
The proof of this theorem, which includes the encoding of a nontrivial property, and a combinatorial argument about the amount of information
needed after a given prefix, appears in an appendix.
The relative expressive power between the first-order temporal logics discussed here
appear in Figure~\ref{fsfr}.

@@

\begin{figure}
\begin{center}
\includegraphics[height=1.1in,width=2.7in]{FIRST.jpg}
\caption{\label{fsfr} An arrow from A to B means that logic  B is more expressive than A}
\end{center}
\end{figure}

@@

Runtime verification of FLTL is performed on an
input that consists of {\em events} in the form of
tuples of relations. In our notation, the input
consists of a sequence ${\cal R}[1] \, {\cal R}[2]  \ldots$,
which we earlier identified with states, where each
${\cal R}[i]$ consists of some relations. A typical use
of runtime verification restricts these relations to have,
per each state just some finite tuples, and, quite commonly, have a single tuple from a single relation per each state.
A set of assignments over the same variables, which has some fixed order
(say alphabetical) also forms a relation.

\klaus{In the text above there are some imprecisions, such as
(1) ``tuples of relations'' - relations are sets of tuples, we don't have tuples of relations. 
(2) ``${\cal R}[i]$ consists of some relations'': it is a set of
relations (I think).
(3) ``to have, per each state just some finite tuples'': tuples are finite. You mean: to contain a finite set of tuples?
(4) what is the story about a set of assignments also being a relation? Needed?
}

\klaus{Below: (\ref{EFLTL}) was not a EPFLTL form of specification.}

Consider again the EPFLTL form
of specification, described in~(\ref{EFLTL}), restricted to the past operators. We now present the definition of EPFLTL, the final main logic proposed in this paper. 
The syntax is as follows:

@@

\[ 
\pi ::= \varphi \mathrm{\ where\ } r_j ( x_j) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \}
\]
\[
\varphi ::= ...
\]
}

@@

%We define $\mathbf{F}$ and $\mathbf{T}$ as special constants
%in order to interpret formulas without free variables.
%They are the complement of each other.
%When combined with set operators (complementation, union, intersection),
%$\mathbf{F}$ behaves as the empty set, hence is idempotent to
%union. Accordingly, $\mathbf{T}$ is idempotent to intersection.
\fi

\klaus{One may of course ask whether the following section on the 
set semantics is needed. Maybe it is needed, since we probably need a semantics. Something to consider.}

>>>>

\begin{figure}
\begin{center}
%\includegraphics[height=0.6in,width=2in]{PROP.jpg}
\includegraphics[width=3.5in]{PROP.jpg}
\caption{\label{prop} An arrow from A to B means that logic  B is more expressive than A. A double headed arrow shows same expressiveness}
\end{center}
\end{figure}

@@

The formulas of the core FLTL logic are 
defined by the following grammar,
where $p$ is a relation,
$a$ is a constant and $x$ is a variable. 
\begin{center}
$\varphi ::= \true  \; | \;
    p ( a ) \; | \;
    p ( x ) \; | \;
    ( \varphi \wedge \varphi ) \;  |   \;
   \neg \varphi \; | \;
   \bigcirc \varphi \; | \; 
   ( \varphi \; \until \; \varphi ) \; | $ %\\
   $ \ominus \; \varphi \; | \;
    ( \varphi  \; \since  \; \varphi ) \; | \;
    \exists x \; \varphi$
\end{center}

@@

Additional operators are defined as in the propositional logic. We define
$\forall x \; \varphi = \neg \exists x \neg \varphi$.
Restricting the modal operators to the past operators
($\since$, $\ominus$ and the ones obtained from them) 
forms the logic PFLTL.

\subsubsection*{Semantics.}
A model is a set of temporal
 relations ${\cal R} = \{ R_1 \, \ldots , R_m \}$.
 Since the standard definition of temporal logic is
 over a sequence (``the execution''), let
 ${\cal R} [ i ] = \{ R_1 [ i ] \, \ldots , R_m [ i ] \}$. ${\cal R} [i]$ represents a {\em state}.
 A model ${\cal R}$ can thus be seen as a sequence
 of states ${\cal R}[1] {\cal R}[2] \ldots$. Let 
 $m$ be a bijection
from relation names (syntax)
onto the relations ${\cal R}$ (semantics).
%After the formal semantical definition, we will abuse notation
%and use the same symbol for the representation of the relation in the logic and for the relation as mathematical object.

@@

Let $\vars ( \varphi )$ be the set of free (i.e., unquantified) variables of a
subformula $\varphi$. 
%The interpretation of a subformula
%$\varphi$ over some finite sequence is as a set of assignments
%to the variables $\vars ( \varphi)$.
We denote by $\gamma |_{\vars ( \varphi )}$ the restriction (projection) of
an assignment $\gamma$ to the free variables appearing in $\varphi$.
%We are careful in combining the sets of assignments
%between different
%subformulas, e.g., through union and intersection (corresponding
%to disjunction and conjunction, respectively) and the quantification.
%This is done using the projection and extension functions over
%assignments defined above.
%
Let $\epsilon$ be the empty assignment (with no variables). 
%
In any of the following cases, $( \gamma,  {\cal R} , i ) \models \varphi$
is defined when $\gamma$ is an
assignment over $\vars ( \varphi )$, and $i\ge 1$.

@@

\item $( \epsilon ,  {\cal R} , i )
 \models p ( a ) $ if $m(p) ( a, i )$, where $a$ denotes a constant from  $d(m(p))$.

\item $( [ x \mapsto a ] , {\cal R} , i ) \models p ( x )$ if $m(p)  ( a , i )$, where $\domain (x) = d ( m ( p ))$.

@@

\item $( \gamma , {\cal R} , i ) \models ( \varphi \; \until \; \psi )$ if
for some $j$, $j \geq i$, $( \gamma |_{\vars ( \psi )} , {\cal R} , j) 
\models \psi $ and for each $k$,  $i \leq k < j $,
$( \gamma |_{\vars ( \varphi )} , {\cal R} , k) \models \varphi$.

@@

For an FLTL (PFLTL) formula with no free variables, denote ${\cal R} \models \varphi$ when $( \epsilon , {\cal R} , 1 ) \models \varphi$.
We will henceforth sometimes abuse notation, as is very often done in logic, and use the same
symbols both for the relations (semantics) and their representation in
the logic (syntax).
The quantification over values of variables here is `static' in the sense that they are independent of the state in the execution. We
denote static quantification with $\exists$ and
$\forall$.
Dynamic quantification, as defined for LTL in Section~\ref{sec:extending-prop-ltl},
and will extend FLTL below, is expressed with $\Exists$ and $\Forall$.

@@

We demonstrate that the lack of expressiveness
carries over from LTL (PLTL) to FLTL (PFLTL).

\vspace{1ex}
\noindent {\bf Example 1.}
Let $p$ and $q$ be temporal relations, where their
restrictions to the state $s_i$, $p[i]$ and $q[i]$, respectively, are unary relations $p$ and $q$. The
specification that we want to monitor is that for each value
$a$, $p(a)$ appears in all the states where $q(a)$ has appeared an even number of times so far (for the odd occurrences, $p(a)$ can also appear, but does not have to). This is an extension of
Wolper's example over LTL, carried over to the first-order case. To show that this is not expressible in FLTL (and PFLTL),
consider models (executions) where only one data element $a$ appears. We can prove by a structural induction on subformulas
that in this case, for each property $\varphi$
we can replace occurrences of variables within
predicates by $a$, i.e., $p(x)$ and $q(x)$ become $p(a)$
and $q  (a )$, respectively, and that we can throw away
the quantification ($\forall x$, $\exists x$), obtaining
an equivalent formula. This is then equivalent to
an LTL formula that is obtained by replacing
the occurrences of $p(a)$ and $q(a)$ by propositions
$p_a$ and $p_b$, which reduces to Wolper's example~\cite{Wolper}.

\vspace{1ex}
\noindent {\bf Example 2.}
Consider the property that asserts that when
$\mathit{inherent}(x , y)$ appears
in a state, denoting that a process 
$y$ wants to inherent variables from a process $x$,
there were a chain of states with process spawning
$\mathit{spawn}(x, x_1)$, $\mathit{spawn} (x_1, x_2)$ $\ldots$
$\mathit{spawn}(x_l , y)$. The reason
is that the required property needs to
assert about the transitive closure
of the relation $\mathit{spawn}$. 
FLTL can be translated in a way similar to a standard translation of $LTL$ to monadic first order logic~\cite{Thomas}. Here, the relations will be written with their explicit last ``time'' parameter, and the temporal operators are replaced with first order quantification.
E.g., $\Box \forall x \, p(x) \rightarrow q(y)$ will be translated into $\forall t \, \forall x \, ((p(x, i) \rightarrow \exists t' p(x, j) )$
\footnote{The obtained model will have, in addition
to the aforementioned relations, also
the linear order $<$ over the naturals.}
It is possible to show that the transitive closure of $\mathit{spawn}$ cannot be expressed in first order setting. The proof is based on the
compactness theory of first order logic~\cite{Flum}.

@@

Extending FLTL (PFLTL) with quantifiers over temporal relations, we obtain QFLTL
(and the past-restricted version QPFLTL, respectively). 
The syntax includes $\Exists p \, \varphi$, 
where $p$ denotes an auxiliary relation.
We also allow $\Forall p\ \varphi = \neg \Exists \neg \varphi$. The semantics is as follows. %We denote by ${\cal R} |_Q$ the set of temporal relations obtained from ${\cal R}$ by removing the set of relations $Q$.

%\klaus{Below, has the notation ${\cal R}' |_{\{ q \} }$ been defined?}

\begin{itemize}
\item $(\gamma ,  {\cal R}, i) \models \Exists q \, \varphi$ iff there exists
${\cal R}'$ such that ${\cal R}' |_{\{ q \} } = {\cal R}$ and 
$( \gamma, {\cal R}' , i) \models \varphi$.
\end{itemize}
Consequently, quantification over relations effectively extends the model ${\cal R}$ into a model ${\cal R}'$ within the scope of the quantifier. Note that quantification here is dynmaic (as in QLTL and QPLTL), since the relations are temporal and can have different sets of tuples in different states.

\subsubsection{Extending FLTL  with Rules.}

We now extend FLTL into EFLTL in a way that is motivated by the  propositional extension from LTL (PLTL) to ELTL (EPLTL). We allow the following formula:
\begin{equation}
\label{EFLTL}
\ \ \ \ \ \psi \mathit{\ where\ } r_j  ( x_j ) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} \mathrm{\ \ \ \     such\ that,}
\end{equation}
\begin{enumerate}
\item $\psi$, the {\em statement}, is an FLTL formula with
no free variables, 
\item $\varphi_j$ are PFLTL formulas with a single
free variable $x_j$,
\item $r_j$ is an auxiliary temporal relation with $2$ parameters; the first parameter is of the same 
type as $x_j$ and the second one is, as usual,
a natural number that is omitted in the temporal formulas. 
An auxiliary relations $r_j$ can appear within $\psi$. They can also appear in $\varphi_k$ of a {\em rule} $r_k := \varphi_k$, but only within the
scope
of a previous-time operator $\ominus$. 
\end{enumerate}

Again, it
is simple to extend the syntax and semantics to include
relations $r_j$ with different number of arguments, and, correspondingly, formulas $\varphi_j$ with corresponding number of free
variables.
The logic EPFLTL is obtained by restricting the temporal modalities of EFLTL to the past ones:
$\since$ and $\ominus$, and those obtained from them.

\begin{lemma} \label{samesame}
The auxiliary temporal relations $r_1, \ldots r_n$ are
uniquely defined from ${\cal R}$.
\end{lemma}
\noindent {\bf Proof.} By a simple induction, similar to Lemma~\ref{fourone}. \qed
\vspace{1ex}
The following formula expresses the property that was shown to be 
not expressible using FLTL, at the beginning of this section.
\begin{equation}
\Box \forall x \, (r(x)\rightarrow p(x)) \mathrm{\ where\ }
r(x) = ( q(x) \leftrightarrow \ominus \neg  r(x)) 
\label{eq:wolper-first-order}
\end{equation}

\noindent
We define\footnote{This is done for lack of space. Formal semantics 
can be given by constructing a set of temporal relations extended
with the auxiliary ones
inductively over growing prefixes, as would be done in a detailed proof of Lemma~\ref{samesame}.} the semantics for the EFLTL 
specification (\ref{EFLTL}) 
by using the following equivalent
QFLTL formula:

@@

\noindent 
Analogously to the propositional case, it is easy to show that EPFLTL does not
need the past temporal operators besides $\ominus$.

%\klaus{Something seems wrong in the above formulation.
%E.g.: Is the $\Box$ a restriction?, what is ``fully quantified single occurrence of some auxiliary relation $r_j$''?}

@@

\begin{theorem}
\label{theo2}
The expressive power of EFLTL is stricktly weaker than that of QFLTL.
\end{theorem}
The proof of this theorem, which includes the encoding of a nontrivial property, and a combinatorial argument, appears in an appendix.
The relative expressive power between the first-order temporal logics presented in this Section
appear in Figure~\ref{fsfr}.

@@

\begin{figure}
\begin{center}
%\includegraphics[height=1.5in,width=3in]{FIRST.jpg}
\includegraphics[width=3in]{FIRST.jpg}
\caption{\label{fsfr} An arrow from A to B means that logic  B is more expressive than A}
\end{center}
\end{figure}

@@

Runtime verification of FLTL is performed on an
input that consists of {\em events} in the form of
tuples. In our notation, the input
consists of a sequence ${\cal R}[1] \, {\cal R}[2]  \ldots$,
which we earlier identified with states, where each
${\cal R}[i]$ consists of the relations in
${\cal R}$,
where the last parameter is restricted to  $i$. A typical use
of runtime verification restricts the
events (tuples) for
each state to a finite number, and often even to a single event.
The RV algorithm will also make
use of sets of assignments over a set of variables for satisfying a subformula at
some state (corresponding to $\old$ and $\current$, which were used with propositional values in Section~\ref{LTLruntime});  these also form a relation over the domains of these variables (having some fixed order).

\ignore{
\klaus{In the text above there are some imprecisions, such as
(1) ``tuples of relations'' - relations are sets of tuples, we don't have tuples of relations. 
(2) ``${\cal R}[i]$ consists of some relations'': it is a set of
relations (I think).
(3) ``to have, per each state just some finite tuples'': tuples are finite. You mean: to contain a finite set of tuples?
(4) what is the story about a set of assignments also being a relation? Needed?
}}

%\subsubsection{Syntax}

\ignore{
\klaus{Below: (\ref{EFLTL}) was not a EPFLTL form of specification.}}

Consider again the EPFLTL form
of specification, described in~(\ref{EFLTL}).

\iffalse
, restricted to the past operators. We now present the definition of EPFLTL, the final main logic proposed in this paper. 
The syntax is as follows:

@@

\[ 
\pi ::= \varphi \mathrm{\ where\ } r_j ( x_j) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \}
\]
\[
\varphi ::= ...
\]}
\fi

@@

\klaus{One may of course ask whether the following section on the 
set semantics is needed. Maybe it is needed, since we probably need a semantics. Something to consider.}
\fi
